---
title: "Comparison of RNA-Seq differential analysis using rnaseqdea"
author: "Wanchang Lin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison of RNA-Seq differential analysis using rnaseqdea}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r common, include=F}
knitr::opts_chunk$set(collapse = FALSE, 
                      messages = FALSE,
                      warning = FALSE,
                      cache = FALSE, 
                      comment = "#>",
                      fig.width = 7, fig.height = 7, 
                      fig.align = "center")
pkgs <- c("mt", "lattice", "latticeExtra", "rnaseqdea", "reshape2")
invisible(lapply(pkgs, library, character.only = TRUE))
```


## Introduction

The R package `rnaseqdea` provides a frame work to call popular RNA-Seq 
differential analysis models like `DESeq2` and `edgeR`. It is easy to 
compare these methods based on identified significant genes, visualisation 
and discrimination. Since the normalisation has impact on the analysis 
results, the normalisation methods provided by `DESeq2` and `edgeR` also
have been implemented to be used by other methods. This vignette uses a 
example data set to show how to compare with different statistical 
analysis methods and normalisation methods.

## Prepare data set

Load the data set: data and meta

```{r }
load(system.file("extdata/vig_data.rda", package = "rnaseqdea"))
```


Filter low expression tags

```{r }
cpms <- t(t(data) / (1e-6 * colSums(data)))
keep <- rowSums(cpms > 0.5) >= 2
sum(keep)
data <- data[keep, ]
```

group information

```{r }
cls <- factor(meta$Treatment)
cls
```

comparison of group 

```{r }
com <- levels(cls)
com
```


## RNA-Seq analysis

We select four modelling methods and four normalisation methods for
comparison.  


```{r }
s.meth <- c("stats_DESeq2", "stats_edgeR", "stats_TSPM", "stats_voom")
n.meth <- c("DESeq", "TMM", "RLE", "UQ")

res <- lapply(s.meth, function(i) {
  cat("\n--Stat = :", i)
  flush.console()
  tmp <- lapply(n.meth, function(j) {
    cat("\n----Norm = :", j)
    flush.console()
    rna_seq_dea(data, cls = cls, com = com, stats.method = i, norm.method = j)
  })
  names(tmp) <- n.meth
  tmp
})
names(res) <- s.meth
```


## Results: Tables

Get stats and rej.num

```{r }
tab <- lapply(res, function(x) lapply(x, function(y) y[c("stats")]))
names(tab) <- gsub("^.*?\\.(.*)", "\\1", names(tab), perl = TRUE)
tab <- un.list(tab)
tab <- lapply(tab, as.data.frame)
```

get reject number

```{r }
rej <- lapply(res, function(x) lapply(x, function(y) y[c("rej.num")]))
rej <- un.list(rej)
names(rej) <- gsub("_rej\\.num", "", names(rej))
names(rej) <- gsub("stats\\.", "", names(rej))
rej <- lapply(rej, as.data.frame)
rej <- do.call("rbind", rej)
```

get 0.05 reject number

```{r }
ind005 <- grep("0\\.05", rownames(rej))
ind001 <- grep("0\\.01", rownames(rej))
ind01 <- grep("0\\.1", rownames(rej))
```

sort rej by threshold

```{r }
rej.ord <- rej[c(ind001, ind005, ind01), ]
rej.ord
```


## Results: Plots

All plots below are based on the normalised data using normal factors.

### PCA plot

```{r }
pca <- lapply(res, function(x) {
  tmp <- lapply(x, function(y) {
    y$pca
  })
  # purrr::compact(tmp)
})

pca <- reshape2::melt(pca, id = c("PC1", "PC2", "cls"))

names(pca) <- c("PC1", "PC2", "cls", "norm", "stat")
pca$stat <- gsub("^.*?\\.(.*)", "\\1", pca$stat, perl = TRUE)

pca.p <-
  useOuterStrips(
    xyplot(PC1 ~ PC2 | stat + norm,
      data = pca, groups = cls, as.table = T,
      xlab = "PC2", ylab = "PC1", main = "PCA Plot",
      auto.key = list(space = "right"),
      par.settings = list(superpose.symbol = list(pch = rep(1:25))),
        panel = function(x, y, ...) {
          panel.xyplot(x, y, ...)
          panel.elli.1(x, y, ...)
        }, ep = 0,
      scales = list(cex = .75, relation = "free")
    )
  )
pca.p
```

### MDS Plot

```{r }
mds <- lapply(res, function(x) {
  tmp <- lapply(x, function(y) {
    y$mds
  })
  # purrr::compact(tmp)
})
mds <- reshape2::melt(mds, id = c("Coord1", "Coord2", "cls"))
names(mds) <- c("Coord1", "Coord2", "cls", "norm", "stat")
mds$stat <- gsub("^.*?\\.(.*)", "\\1", mds$stat, perl = TRUE)

mds.p <-
  useOuterStrips(
    xyplot(Coord1 ~ Coord2 | stat + norm,
      data = mds, groups = cls, as.table = T,
      xlab = "Coordinate 2", ylab = "Coordinate 1", 
      main = "MDS Plot",
      auto.key = list(space = "right"),
      par.settings = list(superpose.symbol = list(pch = rep(1:25))),
        panel = function(x, y, ...) {
          panel.xyplot(x, y, ...)
          panel.elli.1(x, y, ...)
        }, ep = 0,
      scales = list(cex = .75, relation = "free")
    )
  )
mds.p
```

### MA plot

```{r }
ind <- c("mean", "log2.fold.change", "pval", "padj")
tab <- lapply(res, function(x) {
  tmp <- lapply(x, function(y) {
    y$stats[, ind]
  })
})
tab <- reshape2::melt(tab, id = ind) 
names(tab) <- c(ind, "norm", "stat")
tab$stat <- gsub("^.*?\\.(.*)", "\\1", tab$stat, perl = TRUE)

ma.p <-
  useOuterStrips(
    xyplot(log2.fold.change ~ mean | stat + norm,
      data = tab, as.table = T,
      pch = 16, cex = .5,
      col = ifelse(tab$padj < .1, "#FF000040", "black"),
      panel = function(x, y, subscripts, col, ...) {
        panel.xyplot(x, y, col = col[subscripts], ...)
        # panel.loess(x, y, col="green",...)
        # panel.lmline(x, y, col="red",...)
        panel.grid(h = -1, v = 2)
        panel.abline(h = c(-2, 2), col = "blue", lty = 2)
      },
      xlab = "mean", ylab = "log2 fold change",
      main = "MA Plot",
      scales = list(
        x = list(log = TRUE), y = list(log = FALSE, limits = c(-6, 6)),
        cex = .75, relation = "same"
      )
    )
  )
ma.p
```

### Volcano plot

```{r }
v.p <-
  useOuterStrips(
    xyplot(-log10(pval) ~ log2.fold.change | stat + norm,
      data = tab, as.table = T,
      pch = 16, cex = .5,
      col = ifelse(abs(tab$log2.fold.change) > 2 & tab$pval < 0.05, 
                   "#FF000050", "#00000050"),
      panel = function(x, y, subscripts, col, ...) {
        panel.xyplot(x, y, col = col[subscripts], ...)
        panel.grid(h = -1, v = 2)
        panel.abline(v = c(-2, 2), h = -log10(0.05), col = "blue", lty = 2)
      },
      scales = list(cex = .75, relation = "same"),
      main = "Volcano Plot",
      xlab = "log2(Fold Change)", ylab = "-log10(P-value)"
    )
  )
v.p
```

### Histogram of p-values

```{r }
h.p <-
  useOuterStrips(
    histogram(~ pval | stat + norm,
      data = tab, as.table = T,
      type = "count", nint = 100,
      scales = list(cex = .75, relation = "same", y = list(limits = c(0, 30))),
      main = "Histogram of p-values"
    )
  )
h.p
```


## Classification and visualisation by PCA and PLS

Prepare highly expressed data set for classification analysis.


```{r }
## Transform the raw count data by `vst`
data.tr <- vst_rlt_tr(data, method = "vst")  # vst, rlt
dat.list <- lapply(res, function(x) {
  lapply(x, function(y) {
    stats <- y$stats
    sel <- which(stats$pval <= 0.05)
    if (length(sel) != 0) {
      sel <- rownames(stats)[sel]
      dat <- data.tr[sel]
      cls <- cls
      list(dat = dat, cls = cls)
    } else {
      NULL
    }
  })
})

dat.list <- do.call("c", dat.list)
tmp <- gsub("stats\\.", "", names(dat.list))
tmp <- gsub("\\.", "_", tmp)
names(dat.list) <- tmp
```

Call an internal function to perform classification and visualisation by 
PCA and PLS

```{r }
cl <- rnaseqdea:::rna_seq_cl(dat.list, 
  DF = "Comparison", 
  method = c("randomForest", "svm"),
  pars = valipars(sampling = "cv", niter = 20, nreps = 3, strat = TRUE)
)
```

PCA plot

```{r }
cl$pca[[1]]
```

PLS plot

```{r }
cl$pls[[1]]
```

Classification comparison

```{r }
cl$aam.p
```

classification table

```{r }
aam <- reshape2::melt(cl$aam)
aam <- reshape2::dcast(aam, Var1 + L1 ~ Var2, value = "value")
names(aam)[c(1, 2)] <- c("classifier", "algorithm")
aam
```

